---
title: "STAT 6550 HW5"
author: "Blain Morin"
date: "4/12/2021"
output: pdf_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(forecast)
library(knitr)
library(kableExtra)
library(gridExtra)
library(stargazer)

```

In this question, you will analyze a time series of the monthly U.S. production of crude oil (measured in thousands of barrels) from Jan 1980 until Dec 2009. (Source: Energy Information Administration, Dept. of Energy). Our key interest is to understand how oil production has changed over the period of study, by building a time series model that can capture trend, seasonality, and noise. Use R for this question (All the R code you will need can be found on the class website in one place or another).

## (a) Produce a time series plot of the data, and then describe the features of the U.S. oil production series.

```{r}

oil = scan("US_oil_production.txt")
years = seq(from=1980, by=1/12, length=length(oil))
df = data.frame(years = years, oil = oil)

```

```{r}

df %>%
  ggplot(aes(x = years, y = oil)) +
  geom_line() +
  xlab("Years") +
  ylab("Oil (Thousands of Barrels)") +
  ggtitle("Oil Production over Time") +
  scale_x_continuous(breaks = seq(1980, 2010, by = 5)) +
  theme_bw()

```


* There appears to be some dependence or correlation between the points because points that are close to each other or more similar than points that are distant.

* There seems to be a downward trend.

* There appears to be some visual evidence of seasonality, we can see a yearly pattern.

* The variance is mostly constant except for years after 2005.

## (b) Estimate the trend using a moving average of the data calculated over 5 year windows (60 time points). Make sure that you produce estimates of the trend at the start and end of the series. Produce a plot of the estimated trend, superimposed over the original data, and describe it in words.

```{r}

### From CDA Notes

cda.step1 = function (x, d=60)
{
  ## create the filter
  if (d%%2==1) ## d is odd
    our.filter <- rep(1,d)/d
  else ## d is even
    our.filter <- c(0.5, rep(1,d-1), 0.5)/d
  
  n        <- length(x)
  half.d   <- d/2
  longer.x <- c(rep(x[1],half.d), x, rep(x[n],half.d))
  
  ## return the filtered sequence
  stats::filter(longer.x, our.filter)[(half.d+1):(n+half.d)]
}

### Run function,
### Add to our df

df$MA.oil = cda.step1(oil, d = 60)

```

```{r}

df %>%
  ggplot(aes(x = years, y = oil)) +
  geom_line() + 
  geom_line(aes(y = MA.oil), color = "red", size = 1.2, alpha = .7) +
  xlab("Years") +
  ylab("Oil (Thousands of Barrels)") +
  ggtitle("Oil Production over Time (MA(60) In Red)") +
  scale_x_continuous(breaks = seq(1980, 2010, by = 5)) +
  theme_bw()

```

As the red line shows in the above plot, there is a significant decreasing trend through the years 1985 to about 2006. 

## (c) Follow the second step of the classical decomposition algorithm to produce an estimate of the seasonal component (assuming d = 12). Present the estimates in a table (rounded to the nearest 1000 barrels), and produce a graph of these estimates. NOTE: These estimates look unusual! Since there are a different number of days in each month, production will be higher in those months with more days.


```{r}

df$Month = rep(month.name, length(oil)/12)

df = df %>%
  mutate(detrended = oil - MA.oil)

table.c = df %>%
  group_by(Month) %>%
  summarise("Average Barrels (1000s)" = round(mean(detrended), -3)) %>%
  arrange(match(Month, month.name))

kbl(table.c, caption = "Monthly Average Oil Production", booktabs = T) %>%
kable_styling(latex_options = c("striped", "HOLD_position"))

```

```{r}

temp = df %>%
  group_by(Month) %>%
  summarise(mean = round(mean(detrended), -3)) %>%
  arrange(match(Month, month.name))

df %>%
  ggplot(aes(x = years, y = detrended)) +
  geom_line() + 
  geom_hline(aes(yintercept = mean), data = temp, color = "red") +
  facet_wrap(~factor(Month, levels = month.name)) +
  xlab("Year") +
  ylab("Detrended Barrels (1000s)") +
  ggtitle("Estimated Oil Seasonality") +
  theme_bw() +
  theme(strip.background =element_rect(fill="black")) +
  theme(strip.text = element_text(colour = 'white')) 
  

```

## (d) Now calculate the number of 1000 barrels per day. Redo steps (b) and (c) for these per day oil production numbers, showing that the trend is basically the same shape (suitably scaled), but that the pattern of the estimated seasonality is different. Describe the seasonality in words.

```{r}

days.per.month = c(31,28.25,31,30,31,30,31,31,30,31,30,31)
per.day = oil / days.per.month

df$oil = per.day

### From CDA Notes

cda.step1 = function (x, d=60)
{
  ## create the filter
  if (d%%2==1) ## d is odd
    our.filter <- rep(1,d)/d
  else ## d is even
    our.filter <- c(0.5, rep(1,d-1), 0.5)/d
  
  n        <- length(x)
  half.d   <- d/2
  longer.x <- c(rep(x[1],half.d), x, rep(x[n],half.d))
  
  ## return the filtered sequence
  stats::filter(longer.x, our.filter)[(half.d+1):(n+half.d)]
}

### Run function,
### Add to our df

df$MA.oil = cda.step1(per.day, d = 60)

```

```{r}

df %>%
  ggplot(aes(x = years, y = oil)) +
  geom_line() + 
  geom_line(aes(y = MA.oil), color = "red", size = 1.2, alpha = .7) +
  xlab("Years") +
  ylab("Per Day Oil (Thousands of Barrels)") +
  ggtitle("Per Day Oil Production over Time (MA(60) In Red)") +
  scale_x_continuous(breaks = seq(1980, 2010, by = 5)) +
  theme_bw()

```

```{r}

df$Month = rep(month.name, length(oil)/12)

df = df %>%
  mutate(detrended = oil - MA.oil)

table.c = df %>%
  group_by(Month) %>%
  summarise("Average Barrels (1000s)" = round(mean(detrended))) %>%
  arrange(match(Month, month.name))

kbl(table.c, caption = "Monthly Average Oil Production", booktabs = T) %>%
kable_styling(latex_options = c("striped", "HOLD_position"))

```

```{r}

temp = df %>%
  group_by(Month) %>%
  summarise(mean = mean(detrended)) %>%
  arrange(match(Month, month.name))

df %>%
  ggplot(aes(x = years, y = detrended)) +
  geom_line() + 
  geom_hline(aes(yintercept = mean), data = temp, color = "red") +
  facet_wrap(~factor(Month, levels = month.name)) +
  xlab("Year") +
  ylab("Detrended Barrels (1000s)") +
  ggtitle("Estimated Oil Seasonality") +
  theme_bw() +
  theme(strip.background =element_rect(fill="black")) +
  theme(strip.text = element_text(colour = 'white')) 
  

```

* We see that the is more oil production on average in the winter months (from November to May). There is less oil production from June to October. 

## Summarize the detrended and deseasonalized per day oil production series using time series plots, the sample ACF, the sample PACF, and a normal Q-Q plot (Show 60 lags for the ACF and PACF plots). Calculate the Ljung-Box statistic also based on 60 lags. Describe these summaries in words, and pick the most appropriate, but simple, ARMA model for the series, providing reasons for the choice of model that you make.


```{r}

df = df %>%
  group_by(Month) %>%
  mutate(detrend.deseason = detrended - mean(detrended)) %>%
  ungroup()

```

```{r}

# CDA step 3 from class notes

model = lm(detrend.deseason ~ years + I(years^2), data = df)

df = df %>%
  mutate(preds = fitted(model)) %>%
  mutate(resids = resid(model))

a = df %>%
  ggplot(aes(x = years, y = detrend.deseason)) +
  geom_line() +
  geom_line(aes(y = preds), color = "red", size = 1.2, alpha = .8) +
  ggtitle("Deseasonalized and\nDetrended Series") +
  ylab("Barrels Per day (1000s)") +
  xlab("Year") +
  theme_bw()

b = ggAcf(df$resids, lag.max = 60) +
  ggtitle("ACF Plot") + 
  theme_bw() 

c = ggPacf(df$resids, lag.max = 60) +
  ggtitle("PACF Plot") + 
  theme_bw() 

d = df %>%
  ggplot(aes(sample = resids)) +
  stat_qq() +
  stat_qq_line() +
  theme_bw() +
  ggtitle("QQ Plot")

grid.arrange(a, b, c, d, nrow = 2,
             top = "Detrended and Deseasonalized Summary Plots")

```

```{r, echo = TRUE}

Box.test(df$resids, lag = 60, type = "Ljung-Box")

```

## Fit your ARMA model using the arima function. Write down the estimated model, and produce 95% confidence intervals for the model parameters. Interpret your confidence intervals.

```{r, echo = TRUE}

arima16 = arima(df$detrend.deseason,
                order = c(1, 0, 6))


arima16df = data.frame(Est = coef(arima16), SE = sqrt(diag(vcov(arima16))))

temp2 = row.names(arima16df)

arima16df = arima16df %>%
  mutate(Lower = Est - 1.96 * SE) %>%
  mutate(Upper = Est + 1.96 * SE)

arima16df = round(arima16df, 3) %>%
  mutate(Contains0 = ifelse(0>= Lower & 0 <= Upper, "Yes", "No"))

rownames(arima16df) = temp2

kbl(arima16df, booktabs = T, linesep = "", caption = "ARMA (1,6) Coefficients Summary") %>%
kable_styling(latex_options = c("striped", "HOLD_position"))

```

```{r}

df$resid.16 = resid(arima16)

a = df %>%
  ggplot(aes(x = years, y = resid.16)) + 
  geom_line() +
  ylab("ARMA(1,6) Residual") +
  xlab("Year") +
  ggtitle("Residual Plot") +
  theme_bw()

b = ggAcf(df$resid.16, lag.max = 60) +
  ggtitle("ARMA(1,6) ACF Plot") +
  theme_bw()

c = ggPacf(df$resid.16, lag.max = 60) +
  ggtitle("ARMA(1,6) PACF Plot") +
  theme_bw()

d = df %>%
  ggplot(aes(sample = resid.16)) +
  stat_qq() +
  stat_qq_line() +
  theme_bw() +
  ggtitle("QQ Plot")

grid.arrange(a, b, c, d, nrow = 2,
             top = "ARMA (1,6) Diagnostics")
```

```{r}

Box.test(df$resid.16, lag = 60, type = "Ljung-Box")

```


